 

### 2020-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al. 2020; Stark et al. 2020 ), this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

@@

Most notably, light conditions vary  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a, Chazdon & Fetcher 1984; Parker et al. 2019]
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions (Chazdon & Fetcher 1984; Tymen et al. 2017 ; Smith et al. 2019). 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, Doughty and Gouden 2008]

@@

**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al. 2018).
This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; @roberts_stomatal_1990].

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. [@nakamuraForestsTheirCanopies2017, @ozanneBiodiversityMeetsAtmosphere2003]  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

@@

Most notably, light conditions, specifically photosynthetically active radiation (PAR) varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a, Chazdon & Fetcher 1984; Parker et al. 2019]
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984, @tymenQuantifyingMicroenvironmentalVariation2017, @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, @doughtyAreTropicalForests2008]

@@

**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al. 2018).
This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990].
